---
title: "Introduction to RetroJM"
#output: rmarkdown::html_vignette
output: rmarkdown::pdf_document

vignette: >
  %\VignetteIndexEntry{RetroJM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(RetroJM)
```

RetroJM fit a joint model in retrospective time scale, making the longitudinal trajectory starting from death time. It specifically explains how the longitudinal trajectory performs when it's close to death. 

The longitudinal model is: 
$$
Y_i(t^*)=\beta_{\mu}(t^*)+A_i\beta_A(t^*)+X_i^T\psi_X+U_i^*(t^*)+\epsilon_i(t^*)
$$
where

$t^*=D-t$, $D$ is death time and $t$ is observed time.

The hazard for competing risks are: 
$$
\lambda_i^{\delta}(t)=\exp{\Big(\alpha_0^{\delta}(t)+A_i\alpha_A^{\delta}(t)+\alpha_U^{\delta} U_i \Big)},\hspace{0.2cm}\text{for}\hspace{0.12cm} \delta=1,2
$$

## Package installation 

To install, type the following command in R console:

```{r eval=FALSE}
install.packages("RetroJM", repos = "http://cran.us.r-project.org")
```

## Input and Output for RetroJM() function

The `RetroJM()` function is the main function. The User Inputs are: 

- `data`: The dataset used in analysis. It could be either long format or wide format.
It should contain longitudinal measurement, measurement time, a binary variable as
parameter of interest, ID for subjects, event/censoring time, an indicator
for death status, an indicator for drop out status, and any covariates.

- `long_format`:  A logical argument takes value TRUE or FALSE. If TRUE, the input dataset
is in long format. If FALSE, the input dataset should be wide format. Defaut is TRUE.  

- `outcome_var`: Name of outcome variables. If the input data is long format, it should 
be a single character; If the input is wide format, it should be a character vector. 


- `time_var`: Name of observed time variables. If the input data is long format, it should 
be a single character; If the input is wide format, it should be a character vector. It should have same length as "outcome_var". 

- `Ai_var`: Name of parameter of interest. Have to be a binary variable coded as 0/1, 
such as control/treatment.

- `last_obs_time_var`: Name of variable for event/censoring time. 

- `ID_var`: Name of ID variable.

- `surv_censor_var`: Name of indicator variable for death status. 

- `drop_censor_var`: Name of indicator variable for drop out status. 

- `cov_var`: Name of covariates. Default is NULL, meaning no covariates. 

- `longi_end_time`: The end time for longitudinal trajectory. It take numeric as input.
If NULL, then it equals to study end time. Default is NULL.  

- `surv_drop_knots`: The number of additional pieces for piecewise constant hazard model in survival and dropout hazard. Default is 0, meaning no additional pieces and the hazard is constant. 

- `longi_inter_knots`: The number of additional pieces for piecewise linear model in 
longitudinal trajectory. Default is 0, meaning no additional pieces and the longitudinal
trajectory is linear. 

- `num_cores`: The number of cores used for parallel computing. 


## Examples

The example datasets `Retro_data` is included in the package. They could be accessed by: 
```{r}
# library(Retro_JM)

data(Retro_data)
```
The `Retro_data` is a wide format data. It contains 200 subjects, 13 longitudinal measurements. 

From the data, the name of longitudinal measurements are: `Y1-Y13`, name of measured time are: `time1-time13`, name of parameter of interest is: `Ai`, name of event/censoring time is `observed.t`, name of covariates are: `x1, x2`, name of ID variable is: `ID`, name of death status is: `surv_ind`, name of drop out status is: `drop_ind`. We want to stop modeling longitudinal measurement after `25` month after death, and we assume the hazard function has two pieces, longtitudinal measurement has two pieces, then we can fit the model by following: 
```{r, eval=FALSE}
test_run <- RetroJM(
  data = Retro_data,
  long_format = FALSE,
  outcome_var = paste0("Y", 1:13),
  time_var = paste0("time", 1:13),
  Ai_var = "Ai",
  last_obs_time_var = "observed.t",
  cov_var = paste0("x", 1:2),
  ID_var = "ID",
  surv_censor_var = "surv_ind",
  drop_censor_var = "drop_ind",
  longi_end_time = 25,
  surv_drop_knots = 1,
  longi_inter_knots = 1,
  num_cores = 2
)
```

The results could be accessed by: 

```{r, include=FALSE}
load("test_run.Rdata")
```

```{r}
### Estiamtes for survival part
test_run$surv_out_dat

### Estimates for dropout part
test_run$drop_out_dat

### Estimates for longitudinal part
test_run$longiout_dat
```